### 2023-04-04 response-to-audit-report.md

# Response to Audit Report

**TODO:**

- [x] Add prefatory text and executive summary.
- [ ] Add responses and mitigation commits to Isabelle-related findings.
- [x] Add responses and mitigation commits to Haskell-related findings.
- [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
- [ ] Replace broken links and cross references, or use ellipses ("...").
- [ ] Verify the section numbering is correct.
- [ ] Add table of contents.
- [ ] Proofread.

@@

## Executive Summary

Marlowe consists of a domain-specific language (DSL) for the financial industry and for peer-to-peer financial contracts, with a reference semantics written in the Isabelle theorem-proving language and an implementation for the Cardano blockchain written in the Haskell/Plutus programming language. Early in 2023 the Marlowe specifications, on-chain implementation, and validator-test suite were audited by Tweag's High Assurance Software Group: their report is available online and it categorizes findings as either high, medium, or low severity. The present document summarizes the subsequent changes, made in response to the audit's findings, to the Marlowe specifications, its implementation as a Plutus smart-contract validator, and the property-based tests for that validator. All of the high-severity findings have been remedied or mitigated through changes to specifications, proofs, or implementation. Nearly all of the medium- and low-severity ones have also been addressed; justification for not addressing the remainder has been provided.

The high-severity findings related to handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification and the Plutus implementation, and the Isabelle proof of money preservation. Marlowe's Plutus validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states. The fourth finding regarding the Plutus implementation was mitigated to code analysis and property-based testing. The fifth finding about the money-preservation proof was remedied by revising the proof.

The medium-severity findings typically suggested improvements to the Isabelle proofs, addition of more extensive property-based testing, correction of infelicities in the implementation, or safe-coding improvements. The low-severity findings generally centered on the usefulness of code comments, naming of variable bindings, omissions in specifications, and correction of typographical errors. All of the findings that affect the clarity of the specifications, the coverage of testing, or the robustness of implementation were made. Numerous additional property-based tests were added to the test suite as a result of audit findings.

The changes outlined here have not been re-audited. The alterations to Marlowe's Plutus validator (detailed in the Appendix) have been kept to the minimum needed to address the audit findings, but avoiding extensive changes that would might have lessened the relevance of the audit.

Finally, please note that the audit's scope included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts. Although the Marlowe language and its implementation on Cardano aim to make smart contracts safer, it is nevertheless still possible to design individual instances of Marlowe contracts that exhibit unwanted behavior. Such undesirable behavior may result either from poor contract design or lack of testing and analysis of the contract. [A best-practices guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [a security guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) provide advice for the safe and secure design and testing of individual Marlowe contracts, which can themselves be subject to audit or certification.

@@

## Acknowledgement

We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations, and tests and their clear and thorough communication of their findings.

@@

The Marlowe validator had made optimistic assumptions about its own correct operation and hence did not check certain invariants in order to save on Plutus execution costs (item 3 above). The Marlowe semantics validator has now been thoroughly hardened against corruption of initial or final state so that it ensures that the three state-invariants of positive accounts, non-duplication of state entries (accounts, choices, and bound values), and a total internal value matching the script UTxO's value. This hardening was completed without unduly increasing the size (in bytes) of the validator or its Plutus execution cost: such was verified with simulated and on-chain measurement of execution costs for a library of real-life Marlowe contracts. Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator were flawed, at least some of the execution paths of a Marlowe contract remain viable, so funds could be released even under nominally impossible but catastrophic difficulties. The property-based test suite was significantly enhanced to check double-satisfaction situations.

Mitigation of the difference between association lists (item 4 above) in Isabelle (`MList`) and Plutus (`AssocMap`) was handled by the aforementioned enforcement of invariants, by line-by-line code inspection and annotation, and by thoroughly enhanced property-based testing. Work is in progress on a formal proof of the equivalence of `MList` and `AssocMap` under pre-conditions that hold within Marlowe semantics. Note that porting Isabelle's `MList` implementation to Plutus would have enlarged the Marlowe semantics validator by at least 2000 bytes and rendered it too costly to execute within the Cardano ledger's present execution-cost limits.

The audit report for Marlowe on Cardano was based on the commit hash [523f3d56](https://github.com/input-output-hk/marlowe-cardano/commit/523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9). The revisions and mitigations discussed here apply to [ac65eba1](https://github.com/input-output-hk/marlowe-cardano/commit/) of that repository. The appendix to this report list the differences between pre- and post-audit validator code for Marlowe on Cardano.

@@

Although the validator does not use these functions, off-chain code in other Haskell modules does. For the time being, the `contractLifespanUpperBound` and `isClose` have been retained in `Language.Marlowe.Core.V1.Semantics`. They will be relocated in the future if validator and non-validator code is separated into different modules.

### 2.6.6 Comments *(Severity: Low)*

> **File `Semantics.hs`, Function `refundOne`, line *439***
> 
> The comment describing the function is overly concise, as it does not mention the function removing all non-positive accounts before the first positive one, and effectively `uncons`-ing the list.

Commit [ac65eba1](https://github.com/input-output-hk/marlowe-cardano/commit/ac65eba119b4eec8febc3723527edcd99914fdac) elaborates the documentation for `refundOne`.

@@

> **File `Spec/Marlowe/Semantics/Compute.hs`, function `checkFixInterval`, lines *100-102***
> 
> The test `checkFixInterval` is never instantiated with an invalid interval that is in the past, meaning the function `fixInterval` is never tested for that case.

Commit [811f048b](https://github.com/input-output-hk/marlowe-cardano/commit/811f048ba93db10972641c0c235864db1d6014e7) adds a property-based test for an invalid interval in the past.

## Appendix: Post-Audit Changes in Marlowe Validator

```bash
git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 ac65eba119b4eec8febc3723527edcd99914fdac marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
```

### 2023-04-17 response-to-audit-report.md

### 2.1.2 Contracts vulnerable to double satisfaction attacks *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 15`**
> 
> No datum is required for outputs fulfilling payments to addresses generated by the evaluation of a Marlowe contract. This implies that these outputs are vulnerable to double satisfaction in transactions involving other contracts that pay to the same wallets. An example is discussed in […].
> 
> One way to strengthen the implementation is for the Marlowe validator to demand that outputs paid to addresses contain a datum that identifies the contract instance, like the `TxOutRef` of the validator UTxO being spent. Then cooperation with other contracts is possible without double satisfaction if the validators of the other contracts demand a different datum for their outputs.

The Marlowe semantics validator had prevented a limited form of double-satisfaction attack by preventing two Marlowe validators to execute in the same transaction. The audit report correctly highlights that such an attack could be feasible in cases where another Plutus validator (not the selfsame version of the Marlowe semantics validator) executes in a transactions where a payment to an address or to the Marlowe role-payout validator satisfies both the Marlowe semantics validator and the other validator.

Although the audit report's suggestion of including a unique datum for each payment (such as that identifying the contract instance) in any output payment from the contract would likely prevent double satisfaction, implementing such a change in the Marlowe validators would have entailed such extensive changes to the specification and implementation that the overall applicability of the audit findings might subsequently be called into question. Instead, a minimal change to the specification and implementation mitigates this vulnerability: the specification was amended with a "Constraint 20. Single satisfaction" that requires the Marlowe semantics validator be the only Plutus script running in the transaction *if payments to parties are being made in the transaction*. Thus, only allowing one script to validate in case of external payments completely eliminates the possibility that one payment would satisfy two scripts. Other contracts are permitted to cooperate with Marlowe in cases such as deposits, choices, and notifications where Marlowe is not making payments.

### 2023-04-17 response-to-audit-report.md

![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.

### 2.1.12 Different insertion functions used in Isabelle and Haskell code *(Severity: High)*

> **File `Semantics.hs`, Several functions** 
> 
> Where `MList.insert` is used in the Isabelle semantics, [AssocMap.insert](https://github.com/input-output-hk/plutus/blob/v1.0.0/plutus-tx/src/PlutusTx/AssocMap.hs#L147-L148) is used in the Cardano implementation. However, the functions are not equivalent as demonstrated by the following examples: […]
> 
> This renders the Isabelle lemmas inapplicable for the Cardano integration. The lemmas need to demand some properties of an `insert` function without fully spelling it out, or the Cardano integration needs to use `MList.insert` instead of `AssocMap.insert`.
> 
> Similarly, functions `AssocMap.delete` and `MList.delete` differ in behavior when the input map is not sorted: […]
> 
> Functions `AssocMap.lookup` and `MList.lookup` also differ in behavior when the input map is not sorted: […]
> 
> The following usage places were found:
> -   Line 395, `evalValue` depends on `moneyInAccount` which depends on `AssocMap.lookup`.
> -   Line 413, `evalValue` depends on `AssocMap.lookup`.
> -   Line 428, `evalObservation` depends on `AssocMap.member`.
> -   Line 456, function `updateMoneyInAccount` relies on `AssocMap.delete` and `AssocMap.insert`.
> -   Line 482, function `reduceContractStep` relies on `AssocMap.insert`.
> -   Line 567, function `applyAction` relies on `AssocMap.insert`.

The porting of `MList` from Isabelle to Plutus would have increased the size and execution cost of the Marlowe semantics validator beyond the present limits of the Cardano ledger. Work is underway to modify the Isabelle semantics and proofs to make obvious that no assumptions about the implementation of association maps are made. Instead of modifying the Marlowe semantics validator to address this finding, all usages of `AssocMap` were manually reviewed to verify that no vulnerability was introduced by its use (as compared to `MList`) *provided that the precondition of no duplicate entries in the `AssocMap` held*: the source code was annotated with informal reasoning documenting the safety of `AssocMap`'s use. That manual review was augmented by a comprehensive set of property-based tests to check the behavior of `MList` (under the precondition that it was ordered and contained no duplicates) against `AssocMap` (under the precondition that it contained not duplicates) for all functions used in Marlowe semantics or in the Marlowe validator.

@@

Commit [a2ff6aa3](https://github.com/input-output-hk/marlowe-cardano/commit/a2ff6aa334efb7b7955408d29cf0e2edf06bca08) annotates the Marlowe semantics validator with reasoning as to the safe use of `AssocMap`. Commits [af380a29](https://github.com/input-output-hk/marlowe-cardano/commit/af380a292c0369bcc77d8584d81ea80a31326b61), [e2277677](https://github.com/input-output-hk/marlowe-cardano/commit/e2277677987f359dcd8754275eccd5b7c9a880a7), [a95a616a](https://github.com/input-output-hk/marlowe-cardano/commit/a95a616aca59bc7159d3ffa92487f8f021820bff), [b65ccfec](https://github.com/input-output-hk/marlowe-cardano/commit/b65ccfec3708484ff52514671e0e227c13f084f3), and [9e068b6a](https://github.com/input-output-hk/marlowe-cardano/commit/9e068b6a60103f2a40604a7eaea57c5f29b59180) implement the property-based tests for the behavior of the Plutus `AssocMap` against the Isabelle `MList`.

### 2.1.13 Missing specification tests *(Severity: Medium)*

> **File `Spec/Marlowe/Semantics/Compute.hs`** 
> 
> There are no tests for the properties in Section 3 of `specification-v3-rc1.pdf`. Besides checking that there are no translation mistakes, these properties would also help contrasting the assumptions in the Isabelle and the Haskell sides, like the meaning of validity of an association list, which is focused in the previous issue.

Subsequent to the commencement of the audit, a "test oracle" (called test spec) was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification. That generated Haskell implementation is backed by Isabelle proofs of the correctness of Haskell code generation by Isabelle. The test oracle is now applied to the Marlowe semantics implementation in Plutus that the validator use. The test oracle uses the reference implementation to generate test cases to challenge the Plutus implementation and then it checks the Plutus result against its own result. This oracle provides sufficient coverage to address this audit-reports concern regarding the lack of property-based tests for Section 3 of the Marlowe specification.

### 2023-04-17 response-to-audit-report.md

## 2.5 marlowe-cardano specification

### 2.5.1 Lack of guidelines for creating cooperating contracts *(Severity: Medium)*

> **File `marlowe-cardano-specification.md`, Section `Life Cycle of a Marlowe Contract`** 
> 
> Given that transactions are expected to work with Marlowe and non-Marlowe contracts simultaneously, it would be helpful to offer some guidelines for other contracts to avoid double satisfaction. Some degree of cooperation between the contracts that can appear in the same transaction is unavoidable.
> 
> One measure could be to ask every cooperating contract to refrain from paying to the payout validator. In this way, double satisfaction can not affect the payments of the Marlowe contract, if the Marlowe contract only pays to roles rather than addresses.
> 
> Another alternative would be to demand other contracts' outputs to use datums that are different from the roles used by the Marlowe contract for payments.

Commit [00010ea4](https://github.com/input-output-hk/marlowe-cardano/commit/00010ea48439892ff9eec54fb93929f56bcb6c7b) adds a paragraph of guidance to the Marlowe Cardano specification. The mitigation above for [2.1.2](#222-contracts-vulnerable-to-double-satisfaction-attacks-severity-high) implements validator changes to address this.

@@

### 2.5.2 No reference to creating a minting policy *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Monetary Policy for Role Tokens`** 
> 
> The minting policy is not specified, but a reference needs to be offered to explain how to create one.

The monetary policy is not a concern of the Marlowe validator, but commit [fe00af7d](https://github.com/input-output-hk/marlowe-cardano/commit/fe00af7d622bbd17c3327363d4650ee544d1fff7) adds a sentence referencing the discussion of monetary policies for role tokens in the [Marlowe Best Practices Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md).

### 2023-04-17 response-to-audit-report.md

Commit [fda1d3e8](https://github.com/input-output-hk/marlowe-cardano/commit/fda1d3e862fc8f8c31ff479fbd338226e661eb0b) to the Marlowe Cardano specification clarifies that the constraint is supplementary (informational) in nature.

### 2.5.6 Asymmetry between role and wallet payouts *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, `Constraint 15`, **
> 
> The marlowe validator allows multiple outputs to be paid to a wallet, but it demands that a single output exists when paying to a role instead. The motivation to use different approaches needs to be documented. This is implemented in `Scripts.hs` at line 371, in function `payoutConstraints`.

The asymmetry is a convenience and practicality related to the manner in which coin selection and transaction balancing typically occurs nowadays in wallets. Commit [4e81470d](https://github.com/input-output-hk/marlowe-cardano/commit/4e81470de90ddd6858c3ad8aace96e90df5a86ae) adds to the Marlowe Cardano specification a paragraph justifying this asymmetry in payout style and a comment to the Marlowe semantics validator along the same lines.

### 2023-04-17 response-to-audit-report.md

Commit [e150e308](https://github.com/input-output-hk/marlowe-cardano/commit/e150e30867dfa5466639f46bf39656aee2888e38) details the three invariants that the initial state must satisfy: positivity of accounts, non-duplication of state entries (accounts, choices, bound values), and matching the total value in the internal state to the value of the script's UTxO.

### 2.5.9 Unspecified behavior when multiple cases can apply *(Severity: Low)*

> **File `Semantics.hs`, Function `applyCases`, line *597***
> 
> If multiple cases in a case list can apply, the first one is taken. This behavior should be better communicated in the specification.

Although this behavior is required by the Isabelle specification and implemented in the Plutus validator, the comment on `applyCases` has been edited in commit [08d6f34a](https://github.com/input-output-hk/marlowe-cardano/commit/08d6f34ac91da7aa34e7414c5bae0e33e9235d74) to reinforce that the first applicable `Case` is taken for a `When`.

### 2023-04-17 response-to-audit-report.md

## Acknowledgement

We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations, and tests and their clear and thorough communication of their findings.

## Overview of Isabelle Changes

![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here.

@@

Although this `Eq` instance is not used by the Marlowe validators (so it does not have implications for validator security), the equality test has been fixed because it is a liability for off-chain code and for potential future versions of validators that might perform such an equality test. Commit [84d65a70](https://github.com/input-output-hk/marlowe-cardano/commit/84d65a70c29320fab501ad759bbc71855d1b63fc) fixes this and commit [04805a39](https://github.com/input-output-hk/marlowe-cardano/commit/04805a393c0815fe0f2a1a6d7f76d9867ffe0c14) hardens all of the other `Eq` instances to prevent similar problems re-arising if sum types are augmented in the future.

### 2.1.4 Inaccurate formulation of Money preservation *(Severity: High)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
> 
> As the property stands, it is permitted to make deposits in one currency and return payments in a different currency. As long as the sums of the amounts match, the equality is satisfied. Yet it is unlikely that the participants of the contract would agree that money has been preserved.
> 
> Money preservation is a property stated with an equality. The left hand side is the sum of the deposits done by a list of transactions. The right hand side of the equality is the sum of all the payments done in the same list of transactions. Each sum, in turn, is represented as a single integer which aggregates the amounts of the various payments and deposits, irrespective of what currencies correspond to these amounts.

![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.

@@

Commit [d1477610](https://github.com/input-output-hk/marlowe-cardano/commit/d14776101c22eac7fbc67893020f37cbc2a9e339) adds a comment that the use of `foldr` in the Plutus implementation is for efficiency.

### 2.6.10 Constraint implementations differ from description *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Plutus Validator for Marlowe Semantics`** 
> 
> Some constraints mentioned in the specification are written in a different structure than the corresponding constraint in `Scripts.hs`. While such a discrepancy may be useful to minimize verbosity, a unified structure when possible would alleviate a side-by-side comparison. Examples of these differing structures include Constraint 6 and Constraint 14.

The discrepancies of implementation between the constraints in the specification and the Plutus code result from the emphasis on readability in the specification and of efficiency in the implementation. The need to minimize the validator size and execution cost in the implementation has sometimes resulted in less concise or less readable code. Note that the Plutus compiler can be quite sensitive to the precise expression of a constraint as Haskell code.

### 2023-04-19 response-to-audit-report.md

Mitigation of the difference between association lists (item 4 above) in Isabelle (`MList`) and Plutus (`AssocMap`) was handled by the aforementioned enforcement of invariants, by line-by-line code inspection and annotation, and by thoroughly enhanced property-based testing. Work is in progress on a formal proof of the equivalence of `MList` and `AssocMap` under pre-conditions that hold within Marlowe semantics. Note that porting Isabelle's `MList` implementation to Plutus would have enlarged the Marlowe semantics validator by at least 2000 bytes and rendered it too costly to execute within the Cardano ledger's present execution-cost limits.

The audit report for Marlowe on Cardano was based on the commit hash [523f3d56](https://github.com/input-output-hk/marlowe-cardano/commit/523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9). The revisions and mitigations discussed here apply to [ac65eba1](https://github.com/input-output-hk/marlowe-cardano/commit/) of that repository. The appendix to this report list the differences between pre- and post-audit validator code for Marlowe on Cardano.

The medium-severity concerns in the audit report center around missing tests or name shadowing. These have all been remedied either by code changes (in the case of the shadowing) or by the addition of a significant augmentation of the property-based test suite. The audit's low-severity concerns generally relate to insufficiently detailed text in the Marlowe Cardano specification, the need for more elaborate comments in the code, naming of variable bindings, or typographical errors. Nearly all of these finds have been addressed, with only the exception of a few cosmetic recommendations that were not adopted.

@@

## 2.7 Haskell tests

### 2.7.1 More precise failure checks *(Severity: Medium)*

> **File `Spec/Marlowe/Plutus/Specification.hs`, `Various tests`** 
> 
> The tests use the functions `checkSemanticsTransaction` and `checkPayoutTransaction` to verify that various error conditions cause transactions to be rejected. These functions test that a transaction passes or fails, but when it fails, the functions do not consider the error cause. Checking the exact cause is necessary to ensure the transaction is rejected because of the intended reason and not because of some other error condition arising in a particular test case by coincidence.
> 
> The absence of this information makes it easier to accidentally produce a test that is not testing what is intended.

Commit [8d18385b](https://github.com/input-output-hk/marlowe-cardano/commit/8d18385b107fbce2f7a4559e6017d9e228ab958b) to the Marlowe test suite for `checkSemanticsTransaction` instruments Plutus scripts for the Marlowe validator so that the precise cause of the test failure is verified.
